# Which One to Use: for or the purrr map family functions?

## Functionals in R: A simple introduction to map family functions in `purrr`



```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "What's wrong with for loops? Nothing. (AI generated image based on author prompt)"
library(knitr)
library(knitr)
library(gridExtra)
library(grid)

p1 <- rasterGrob(png::readPNG("purrr.png"))
p2 <- rasterGrob(png::readPNG("purrrremove.png"))

grid.arrange(p1, p2, ncol = 2)
```

When we start programming in R, one of the first things we learn is the for loop. A for loop allows us to repeat an operation multiple times.

For example, suppose we want to calculate the square of the numbers from 1 to 5, then we can write a simple `for` loop as follows:

```{r}
result <- numeric(5)

for (i in 1:5) {
  result[i] <- i^2
}
result
```

There is another way to express the same idea using **functionals**.

## What is a functional?

A functional is a function that takes another function as an input and uses that function to perform an operation.

Here is an example code to illustrate the idea of functionals. 

```{r}
apply_function <- function(x, f) {
  f(x)
}
```

The `apply_function` takes the function `f` as an input to the function. In other words `f` is a function passed as an argument. Now we can pass different functions to the `apply_function` as follows:

```{r}
apply_function(5, sqrt)
```

or

```{r}
apply_function(5, abs)
```

or you can pass your own function as the input to the function as illustrated below.

```{r}
power <- function(x, n) {
  x^n
}

apply_function_multi_input <- function(x, n, f) {
  f(x, n)
}

apply_function_multi_input(5, 2, power)

```


## `map` family functions in the purrr package

The purrr package in the tidyverse provides many useful functionals. Some of the most commonly used illustrated below. These map() family functions are particularly useful when we want to apply the same operation repeatedly.

### `map()`: apply a function repeatedly

Suppose we want to square the numbers from 1 to 5.

```{r}
library(purrr)
map(1:5, function(x) {
  x^2
})
```

The important part is:

```{r}
function(x) {
  x^2
}
```

This is the function that we want to apply. The `map()` function takes each value in 1:5 and gives it to this function. The output is stored as a list.

Why does map() return a list?: A list is flexible because the result of each iteration does not necessarily have to be the same type. 

Here is another example:

```{r}
map(1:3, function(x) { data.frame( number = x, square = x^2 ) })
```

### `map_dbl()`: when the result is numeric

If we know that every iteration produces a single numeric value, we can use map_dbl().

```{r}
map_dbl(1:5, function(x) {
  x^2
})
```

The result is a numeric vector.

### `map_dfr()`: when each iteration produces a data frame

Now the outputs from each iteration is stored in dataframe. Here, `map_dfr()` applies the function three times and then combines the resulting data frames by rows.

This is particularly useful when working with data analysis problems where every iteration generates a data frame with a single row that we want to combine into one larger data frame.


```{r}
map_dfr(1:3, function(i) {
  data.frame(
    number = i,
    square = i^2
  )
})
```

### Other map family functions

The map() family has different versions depending on the type of output you want:

```{r}
library(tibble)

functional_table <- tribble(
  ~Functional,   ~Output,                         ~Example,
  "map()",       "List",                          "map(1:5, square)",
  "map_dbl()",   "Numeric / double",              "map_dbl(1:5, square)",
  "map_int()",   "Integer",                       "map_int(1:5, as.integer)",
  "map_chr()",   "Character",                     "map_chr(1:5, as.character)",
  "map_lgl()",   "Logical (TRUE/FALSE)",          "map_lgl(1:5, function(x) x > 3)",
  "map_dfr()",   "Data frame (rows combined)",    "map_dfr(1:5, function(x) data.frame(x))"
)

functional_table
```


## Functionals and your own functions

We can define our own function and pass into map family functions as well. 

For example,

```{r}
square <- function(x) {
  x^2
}
```

Then:

```{r}
map_dbl(1:5, square)
```

## What is wrong with a for loop? 

There is nothing wrong with a `for` loop. In fact, a for loop and a functional are often two different ways of expressing the same computational idea. Functionals are not simply a shorter replacement for `for` loops.

With a `for` loop, we tell R how to repeat an operation.

With a `purrr` functional, we tell R what operation we want, and purrr handles the repetition.

In simple terms:

for says: “Repeat this.”

purrr says: “Do this to each one.”

In otherwords, with purrr, we instead of repeatedly writing the mechanism for iteration, describe the operation you want to perform and **let the functional handle the repetition**.

For example

```{r}
# for loop
result <- numeric(5)

for (i in 1:5) {
  result[i] <- i^2
}
```

With purrr:

```{r}
map_dbl(1:5, function(x) x^2)
```

The purrr version focuses on the operation. It says, "square each number". While `map_dbl()` takes care of the iteration.

## Which one to use? purrr or for

There isn't a rule that says always use for or always use purrr. Both are useful; the choice depends on the task and what makes the code clearest.

Use `purrr` when,

  - The operation is simple and repetitive
  
  - You want to apply a function to each element
  
  - You want concise, readable functional code
  
  - You want to control the output type
  
  - You are working with lists or nested data
  
  - You are already comfortable with tidyverse
    
    
